---
title: "FecDist pipeline: QC report"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(ggplot2)
library(DT)

theme_set(theme_bw())
type_colours <- c(drones = "#1b9e77", workers = "#d95f02")

# Paths, relative to results/qc/ where the pipeline renders a copy of this file
qc_dir <- "."
bqsr_dir <- file.path("..", "bqsr")

# Amel_HAv3.1: the 16 chromosomes (labelled 1-16) and the mitochondrion
chromosomes <- sprintf("NC_0376%02d.1", 38:53)
chromosome_labels <- setNames(as.character(1:16), chromosomes)
mitochondrion <- "NC_001566.1"

# Thresholds to flag pools: none yet

samples <- read_tsv(file.path(qc_dir, "samples.tsv"), show_col_types = FALSE)
drones <- samples$sample[samples$type == "drones"]

show_table <- function(df) {
  datatable(df, rownames = FALSE, filter = "top",
            options = list(pageLength = 25, scrollX = TRUE))
}
```

```{r load-samtools}
# samtools stats: one "TAG<tab>field..." line per value; SN = summary numbers,
# IS = insert sizes, FFQ/LFQ = quality counts per cycle of first/last reads
stats_lines <- function(lines, tag) {
  strsplit(lines[startsWith(lines, paste0(tag, "\t"))], "\t")
}

mean_quality_per_cycle <- function(fields, read) {
  map_dfr(fields, function(f) {
    counts <- as.numeric(f[-(1:2)])
    tibble(read = read, cycle = as.numeric(f[2]),
           mean_quality = sum((seq_along(counts) - 1) * counts) / sum(counts))
  })
}

read_samtools_stats <- function(sample) {
  lines <- readLines(file.path(qc_dir, "samtools_stats", paste0(sample, ".stats")))
  list(
    sn = map_dfr(stats_lines(lines, "SN"), function(f) {
      tibble(key = sub(":$", "", f[2]), value = as.numeric(f[3]))
    }) |> mutate(sample = sample),
    is = map_dfr(stats_lines(lines, "IS"), function(f) {
      tibble(insert_size = as.numeric(f[2]), pairs = as.numeric(f[3]))
    }) |> mutate(sample = sample),
    quality = bind_rows(
      mean_quality_per_cycle(stats_lines(lines, "FFQ"), "R1"),
      mean_quality_per_cycle(stats_lines(lines, "LFQ"), "R2")
    ) |> mutate(sample = sample)
  )
}

stats <- map(samples$sample, read_samtools_stats)

mapping <- map_dfr(stats, "sn") |>
  pivot_wider(names_from = key, values_from = value) |>
  transmute(
    sample,
    reads_M = `raw total sequences` / 1e6,
    mapped_pct = 100 * `reads mapped` / `raw total sequences`,
    properly_paired_pct = 100 * `reads properly paired` / `raw total sequences`,
    duplicates_pct = 100 * `reads duplicated` / `raw total sequences`,
    mapq0_pct = 100 * `reads MQ0` / `reads mapped`,
    error_rate = `error rate`,
    insert_size_mean = `insert size average`,
    insert_size_sd = `insert size standard deviation`
  )
```

```{r load-mosdepth}
# Per contig mean depth (rows "<contig>_region" repeat the windows: dropped)
depth_summary <- map_dfr(samples$sample, function(s) {
  read_tsv(file.path(qc_dir, "mosdepth", paste0(s, ".mosdepth.summary.txt")),
           show_col_types = FALSE) |>
    filter(!grepl("_region$", chrom)) |>
    mutate(sample = s)
})

# Cumulative distribution: fraction of bases covered at >= depth (whole genome)
depth_dist <- map_dfr(samples$sample, function(s) {
  read_tsv(file.path(qc_dir, "mosdepth", paste0(s, ".mosdepth.global.dist.txt")),
           col_names = c("chrom", "depth", "fraction"), show_col_types = FALSE) |>
    filter(chrom == "total") |>
    mutate(sample = s)
})

depth_windows <- map_dfr(samples$sample, function(s) {
  read_tsv(file.path(qc_dir, "mosdepth", paste0(s, ".regions.bed.gz")),
           col_names = c("chrom", "start", "end", "depth"), show_col_types = FALSE) |>
    filter(chrom %in% chromosomes) |>
    mutate(sample = s)
})

# Mean depth over the 16 chromosomes, used to normalise
chromosome_depth <- depth_summary |>
  filter(chrom %in% chromosomes) |>
  group_by(sample) |>
  summarise(mean_depth = sum(bases) / sum(length))

fraction_at <- function(s, x) {
  d <- depth_dist |> filter(sample == s, depth == x)
  if (nrow(d) == 0) 0 else d$fraction
}

depth <- chromosome_depth |>
  left_join(depth_summary |> filter(chrom == mitochondrion) |>
              select(sample, mito_depth = mean), by = "sample") |>
  mutate(
    mito_ratio = mito_depth / mean_depth,
    ge10x_pct = 100 * map_dbl(sample, fraction_at, 10),
    ge20x_pct = 100 * map_dbl(sample, fraction_at, 20),
    ge50x_pct = 100 * map_dbl(sample, fraction_at, 50),
    ge100x_pct = 100 * map_dbl(sample, fraction_at, 100)
  ) |>
  select(-mito_depth)
```

```{r load-bqsr}
# GATK recalibration report: several tables, each "#:GATKTable:<name>:",
# a header line, whitespace separated rows, and a blank line
read_gatk_table <- function(file, name) {
  lines <- readLines(file)
  start <- which(lines == paste0("#:GATKTable:", name, ":"))
  end <- start + 1
  while (end < length(lines) && nzchar(trimws(lines[end + 1]))) end <- end + 1
  fields <- strsplit(trimws(lines[(start + 1):end]), "\\s+")
  table <- as_tibble(do.call(rbind, fields[-1]), .name_repair = "minimal")
  names(table) <- fields[[1]]
  type_convert(table, col_types = cols())
}

# round<r>.table measures the BAM entering round r (round1 = before any
# recalibration); final.table measures the final BAM
bqsr_files <- map_dfr(samples$sample, function(s) {
  files <- list.files(file.path(bqsr_dir, s), pattern = "^(round[0-9]+|final)\\.table$",
                      full.names = TRUE)
  tibble(sample = s, round = sub("\\.table$", "", basename(files)), file = files)
})
has_bqsr <- nrow(bqsr_files) > 0
rounds <- unique(bqsr_files$round)
round_levels <- c(sort(rounds[rounds != "final"]), intersect("final", rounds))
bqsr_files <- mutate(bqsr_files, round = factor(round, round_levels))

read_bqsr <- function(name) {
  pmap_dfr(bqsr_files, function(sample, round, file) {
    read_gatk_table(file, name) |> mutate(sample = sample, round = round)
  }) |>
    filter(EventType == "M") |>
    rename(lane = ReadGroup)
}

bqsr_overall <- tibble(sample = character(), lane = character(), round = factor(),
                       reported = numeric(), empirical = numeric(), difference = numeric())
if (has_bqsr) {
  bqsr_overall <- read_bqsr("RecalTable0") |>
    transmute(sample, lane, round, reported = EstimatedQReported,
              empirical = EmpiricalQuality, difference = reported - empirical)
  bqsr_by_quality <- read_bqsr("RecalTable1")
  bqsr_by_cycle <- read_bqsr("RecalTable2") |>
    filter(CovariateName == "Cycle") |>
    mutate(cycle = as.numeric(CovariateValue)) |>
    group_by(sample, lane, round, cycle) |>
    summarise(observations = sum(Observations), errors = sum(Errors), .groups = "drop") |>
    mutate(
      read = if_else(cycle > 0, "R1", "R2"),
      cycle = abs(cycle),
      empirical = -10 * log10((errors + 1) / (observations + 2))
    )
}
```

```{r load-allele-balance}
allele_balance <- tibble(sample = character(), het_median = numeric(),
                         het_below_0.2_pct = numeric())

if (length(drones) > 0) {
  ab_counts <- map_dfr(drones, function(s) {
    read_tsv(file.path(qc_dir, "allele_balance", paste0(s, ".summary.tsv")),
             show_col_types = FALSE)
  })

  ab_sites <- map_dfr(drones, function(s) {
    read_tsv(file.path(qc_dir, "allele_balance", paste0(s, ".sites.tsv.gz")),
             show_col_types = FALSE) |>
      mutate(sample = s)
  }) |>
    mutate(depth = ref_ad + alt_ad, fraction = alt_ad / depth)

  allele_balance <- ab_sites |>
    group_by(sample) |>
    summarise(
      het_median = median(fraction[genotype == "het"]),
      het_q25 = quantile(fraction[genotype == "het"], 0.25),
      het_q75 = quantile(fraction[genotype == "het"], 0.75),
      het_below_0.2_pct = 100 * mean(fraction[genotype == "het"] < 0.2),
      het_above_0.8_pct = 100 * mean(fraction[genotype == "het"] > 0.8),
      hom_median = median(fraction[genotype == "hom"]),
      hom_below_0.9_pct = 100 * mean(fraction[genotype == "hom"] < 0.9)
    )
}
```

## Summary

One row per pool. Mapping from samtools stats, depth from mosdepth (over the 16
chromosomes), allele balance for drone pools only.

```{r summary}
samples |>
  select(sample, type) |>
  left_join(mapping |> select(sample, reads_M, mapped_pct, duplicates_pct), by = "sample") |>
  left_join(depth |> select(sample, mean_depth, mito_ratio), by = "sample") |>
  left_join(allele_balance |> select(sample, het_median, het_below_0.2_pct),
            by = "sample") |>
  mutate(across(where(is.double), \(x) round(x, 2))) |>
  show_table()
```

## Mapping

samtools stats on the final BAM of each pool. Base qualities in this BAM are
already recalibrated by BQSR.

```{r mapping-table}
mapping |> mutate(across(where(is.double), \(x) signif(x, 4))) |> show_table()
```

### Insert size

Fraction of read pairs per insert size, one line per pool.

```{r insert-size}
map_dfr(stats, "is") |>
  group_by(sample) |>
  mutate(fraction = pairs / sum(pairs)) |>
  ungroup() |>
  filter(insert_size <= 1000) |>
  left_join(samples, by = "sample") |>
  ggplot(aes(insert_size, fraction, group = sample, colour = type)) +
  geom_line(alpha = 0.4) +
  scale_colour_manual(values = type_colours) +
  labs(x = "Insert size (bp)", y = "Fraction of pairs", colour = NULL)
```

### Base quality along reads

Mean base quality per cycle (recalibrated), one line per pool.

```{r quality-per-cycle}
map_dfr(stats, "quality") |>
  left_join(samples, by = "sample") |>
  ggplot(aes(cycle, mean_quality, group = sample, colour = type)) +
  geom_line(alpha = 0.4) +
  facet_wrap(~read) +
  scale_colour_manual(values = type_colours) +
  labs(x = "Cycle", y = "Mean base quality", colour = NULL)
```

## Depth

mosdepth on the final BAM: reads with MAPQ >= 15, duplicates excluded. Mean
depth over the 16 chromosomes; `mito_ratio` = mitochondrion / chromosome depth;
`geNx_pct` = % of the whole genome covered at >= N x.

```{r depth-table}
depth |> mutate(across(where(is.double), \(x) signif(x, 4))) |> show_table()
```

### Cumulative depth

Fraction of the genome covered at >= depth, up to 3 x each pool's mean depth.

```{r depth-cumulative}
depth_dist |>
  left_join(chromosome_depth, by = "sample") |>
  filter(depth <= 3 * mean_depth) |>
  left_join(samples, by = "sample") |>
  ggplot(aes(depth, fraction, group = sample, colour = type)) +
  geom_line(alpha = 0.4) +
  facet_wrap(~type, scales = "free_x") +
  scale_colour_manual(values = type_colours, guide = "none") +
  labs(x = "Depth", y = "Fraction of genome at >= depth")
```

### Depth per chromosome

Mean depth of each chromosome divided by the pool's mean depth (1 = average).

```{r depth-per-chromosome}
#| fig-height: !expr max(4, 0.15 * nrow(samples))
depth_summary |>
  filter(chrom %in% chromosomes) |>
  left_join(chromosome_depth, by = "sample") |>
  mutate(relative = mean / mean_depth,
         chromosome = factor(chromosome_labels[chrom], chromosome_labels)) |>
  ggplot(aes(chromosome, sample, fill = relative)) +
  geom_tile() +
  scale_fill_gradient2(midpoint = 1, low = "#2166ac", high = "#b2182b") +
  labs(x = "Chromosome", y = NULL, fill = "Relative\ndepth")
```

### Depth along chromosomes

Mean depth in windows, divided by the pool's mean depth. Line: median over the
pools of each type; ribbon: 25-75 % of the pools.

```{r depth-along-chromosomes}
#| fig-height: 10
depth_windows |>
  left_join(chromosome_depth, by = "sample") |>
  left_join(samples, by = "sample") |>
  mutate(relative = depth / mean_depth,
         position_Mb = (start + end) / 2e6,
         chromosome = factor(chromosome_labels[chrom], chromosome_labels)) |>
  group_by(type, chromosome, position_Mb) |>
  summarise(median = median(relative), q25 = quantile(relative, 0.25),
            q75 = quantile(relative, 0.75), .groups = "drop") |>
  ggplot(aes(position_Mb, median, colour = type, fill = type)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~chromosome, scales = "free_x", ncol = 4) +
  scale_x_continuous(breaks = seq(0, 30, by = 5)) +
  scale_colour_manual(values = type_colours) +
  scale_fill_manual(values = type_colours) +
  coord_cartesian(ylim = c(0, 2.5)) +
  labs(x = "Position (Mb)", y = "Relative depth", colour = NULL, fill = NULL)
```

## Base quality recalibration

`round<r>` tables measure the BAM entering round r (`round1`: qualities from
the sequencer), `final` measures the final BAM, with the last round's known
sites. After recalibration, reported and measured (empirical) qualities should
agree.

```{r bqsr-table}
#| eval: !expr has_bqsr
bqsr_overall |> mutate(across(where(is.double), \(x) round(x, 2))) |> show_table()
```


### Measured vs reported quality

One line per lane; dashed: perfect agreement. Quality bins with fewer than
1000 bases are left out.

```{r bqsr-by-quality}
#| eval: !expr has_bqsr
bqsr_by_quality |>
  filter(Observations >= 1000) |>
  ggplot(aes(QualityScore, EmpiricalQuality, group = interaction(sample, lane))) +
  geom_abline(linetype = "dashed") +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3, size = 0.8) +
  facet_wrap(~round) +
  labs(x = "Reported quality", y = "Measured quality")
```

### Measured quality along reads

One line per lane.

```{r bqsr-by-cycle}
#| eval: !expr has_bqsr
#| fig-height: 7
bqsr_by_cycle |>
  ggplot(aes(cycle, empirical, group = interaction(sample, lane))) +
  geom_line(alpha = 0.3) +
  facet_grid(round ~ read) +
  labs(x = "Cycle", y = "Measured quality")
```

## Drone pools: allele balance

The drones of a pool are brothers, so every SNP is carried by 0, 1 or 2 of the
queen's copies: expected alternate fraction around 50 % at `0/1` sites and
around 100 % at `1/1` sites. A second peak at low fractions at `0/1` sites
points to foreign drones.


Sites: SNPs called `0/1` or `1/1` in the gVCF, depth >= `min_depth` and <=
`max_depth` (4 x the pool's median depth), random sample of at most 50,000 SNPs per
genotype.

```{r ab-counts}
#| eval: !expr length(drones) > 0
ab_counts |> show_table()
```

```{r ab-table}
#| eval: !expr length(drones) > 0
allele_balance |> mutate(across(where(is.double), \(x) round(x, 3))) |> show_table()
```

### Alternate fraction at `0/1` sites

One row per pool, bins of 2 %.

```{r ab-het-heatmap}
#| eval: !expr length(drones) > 0
#| fig-height: !expr max(4, 0.15 * length(drones))
ab_sites |>
  filter(genotype == "het") |>
  mutate(bin = pmin(floor(fraction / 0.02), 49) * 0.02 + 0.01) |>
  count(sample, bin) |>
  group_by(sample) |>
  mutate(share = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(bin, sample, fill = share)) +
  geom_tile() +
  scale_fill_viridis_c() +
  coord_cartesian(xlim = c(0, 1)) +
  labs(x = "Alternate allele fraction", y = NULL, fill = "Share\nof sites")
```

### Alternate fraction at `1/1` sites

One line per pool.

```{r ab-hom}
#| eval: !expr length(drones) > 0
ab_sites |>
  filter(genotype == "hom") |>
  ggplot(aes(fraction, group = sample)) +
  geom_density(colour = scales::alpha("black", 0.3)) +
  coord_cartesian(xlim = c(0, 1)) +
  labs(x = "Alternate allele fraction", y = "Density")
```
